D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-01-30 02language_design.tex
alg moved to sec 3

These equational axioms provide an operable theory for the proof automation algorithm. Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
\begin{lemma}
    For all well-formed contexts \( E \) and terms \( e_1, e_2 \), if \( E \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
\end{lemma}
\begin{proof}
    This follows directly from checking all cases.
\end{proof}

With these syntax, typing and semantics introduced, we demonstrate the formalization of the symbols used in the motivating example to conclude this section.
\begin{example}[Formalizing the Motivating Example]
    \label{ex: formalizing motivating}
    Definitions and assumptions in the environment \( E \) are formalized as follows:
    \begin{align*}
        & \text{TPO} && := \lambda T_1 : \Index. \lambda T_2 : \Index. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
        &\text{phi} &&:= \lambda T : \Index. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
        & T && : \Index; \\
        & M && : \OType(T, T).
    \end{align*}
    Notice how the functions and higher-order typing helps to formalize the abstract concepts here.
    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \). And its type \(\forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1)\) suggests that TPO is a function from operators to operators, depending on the index $T_1$ and $T_2$.
    The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the plain Dirac notation as:
    \[
    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
    \]
\end{example}

>>>>

These equational axioms provide an operable theory for the proof automation algorithm. Denotational semantics can be seen as one model for this theory, meaning that equivalences derived from the axioms always imply equivalence in the interpretations.
\begin{lemma}
    \label{lem: axiom sound}
    For all well-formed contexts \( E \) and terms \( e_1, e_2 \), if \( E \vdash e_1 = e_2 \), then \( \sem{e_1} = \sem{e_2} \).
\end{lemma}
\begin{proof}
    This follows directly from checking all cases.
\end{proof}

With these syntax, typing and semantics introduced, we can formalize~\Cref{eq: motivating plain}, the motivating~\Cref{ex: motivating} in plain Dirac notation.
\begin{example}[Motivating Example Formalization]
    \label{ex: formalizing motivating}
    Definitions and assumptions in the context \( E \) are formalized as follows:
    \begin{align*}
        & \text{TPO} && := \lambda T_1 : \Index. \lambda T_2 : \Index. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
        &\text{phi} &&:= \lambda T : \Index. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
        & T && : \Index; \\
        & M && : \OType(T, T).
    \end{align*}
    Notice how the functions and higher-order typing helps to formalize the abstract concepts here.
    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \). And its type \(\forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1)\) suggests that TPO is a function from operators to operators, depending on the index $T_1$ and $T_2$.
    The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the plain Dirac notation as:
    \[
    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
    \]
\end{example}

\subsection{Algorithm to Decide Equivalence}

The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax. We use an efficient algorithm to perform the normalization totally on $R \cup F$.

\begin{enumerate}
    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
    \item \textbf{Variable Expansion}: Convert to abstract element-wise representation.
    \item \textbf{Second Rewritings}: Normalize terms on \( R \) modulo \( F \).
    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalence.
    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
\end{enumerate}

Here steps 1 - 3 are term rewriting for $R$. 
Term rewriting reduces a term repeatedly using a set of rules in the form of $l\ \reduce\ r$. The reduction proceeds by matching the subterms with the left-hand side, and replacing it with the right-hand side. 
For example, the rule $a.K + K\ \reduce\ (a + 1).K$ matches the term $(x\times y) . \ket{t} + \ket{t}$ by the substitution $\{a \to  x\times y, K \to \ket{t}\}$, and is rewritten into $(x\times y + 1) . \ket{t}$.
Step 1 and 3 uses the same set of rewriting rules, see~\Cref{sec: rewriting rules}.
Step 2 expands variables like the ket $K$ to their abstract element-wise representation $\sum_{i} (\bra{i} \cdot K). \ket{i}$, which improves the decidability with sum.

Steps 4 - 6 are special algorithms to further normalize the axiom set $F$.
Here the main difficulty is the coexistence of AC-equivalence and SUM-SWAP, so that naive sorting will not convert equivalent terms to the same form.
For step 4 and 5, the key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted. Therefore we first ignore bound variables and normalize AC-equivalence by sorting. Then the order of summation can be decided accordingly. The last step converts functions to de Bruijn indices~\cite{deBruijn1972lambda} to decide $\alpha$-equivalence. For detailed descriptions, see~\Cref{sec: decide}.

As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $F$ by iterating through all possible permutations.
Our algorithm totally normalize the term, as is illustrated below.
\[
\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
\qquad
\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
\]
While the permutation checking algorithm has factorial complexity on the number of AC symbol arguments, our algorithm uses normalization by sorting, and the complexity is polynomial on the term size. This improvement in efficiency is also confirmed by experiments.

% Lastly, we can prove that the equivalence established by this normalization procedure is sound with respect to the semantics.

% \begin{theorem}[Soundness]
%     For any well-formed context \( E \) and well-typed expressions \( e_1 \) and \( e_2 \), if  $e_1$ and $e_2$ have the same normal form, then \( \sem{e_1} = \sem{e_2} \).
% \end{theorem}
% \begin{proof}
%     Because of~\Cref{lem: axiom sound}, we only need to prove soundness with respect to axioms. 
%     This is because the term rewriting procedure follows from the fact that each rewriting rule preserves equivalence. Furthermore, the operations in the sort and swap transformations respect the AC-equivalence and SUM-SWAP axioms. Finally, the de Bruijn normalization ensures soundness for \( \alpha \)-equivalence.
% \end{proof}

### 2025-01-30 05tool_eval.tex
alg moved to sec 3

\section{Implementation and Case Study}
The refinements and extensions above concludes in our implementation called D-Hammer, a solver written in \CC. It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel. The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output. 
It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
This implementation is tested on the benchmark of the previous work and new examples for labelled Dirac notation, demonstrating significant improvement in expressivity and efficiency.

@@

\subsection{Equivalence Checking Algorithm}

The equivalence of Dirac notations are decided by normalization, which transforms equivalent expressions under a set of axioms into the same syntax. We use an efficient algorithm to perform the normalization totally on $R \cup F$.

\begin{enumerate}
    \item \textbf{First Rewritings}: Expand definitions and simplify expressions.
    \item \textbf{Variable Expansion}: Convert to abstract element-wise representation.
    \item \textbf{Second Rewritings}: Normalize terms on \( R \) modulo \( F \).
    \item \textbf{Sorting Without Bound Variables}: Normalize AC-equivalence.
    \item \textbf{Swap Successive Summations}: Normalize SUM-SWAP equivalence.
    \item \textbf{De Bruijn Normalization}: Normalize \( \alpha \)-equivalence.
\end{enumerate}
Here steps 1 - 3 are term rewriting for $R$, and steps 4 - 6 are special algorithms to further normalize the axiom set $F$. Step 2 expands variables like $K$ to their abstract element-wise representation $\sum_{i} (\bra{i} \cdot K). \ket{i}$, which improves the decidability with sum. For step 4 and 5, the key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted. Therefore we first ignore bound variables and normalize AC-equivalence by sorting. Then decide the order of summation accordingly. The last step converts functions to de Bruijn indeices~\cite{deBruijn1972lambda} to decide $\alpha$-equivalence.

As a comparison, the previous work conducts normalization paritally on $R$, and decide axioms in $F$ by iterating through all possible permutations.
Our algorithm totally normalize the term and significantly improves the efficiency.
\[
\text{(partial)}\qquad e_1 \mathop{\twoheadrightarrow}^R e_1' \mathop{=}^F e_2' \mathop{\twoheadleftarrow}^R e_2
\qquad
\text{(total)}\qquad e_1 \mathop{\twoheadrightarrow}^{R \cup F} e \equiv e \mathop{\twoheadleftarrow}^{R \cup F} e_2
\]

\yx{we can add a normalization example here.}

\subsection{Use Case}
Here we encode the motivating~\Cref{ex: motivating}, examine and explain how to check it using D-Hammer. The encoding is shown below.

@@

{
    \center
    \begin{tabular}{c|c c c|c c c}
        \hline
        \multirow{2}{*}{source} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{D-Hammer} \\
        \cline{2-7}
                                 & expressable & success & time(s)           & expressable & success & time(s)                 \\
        \hline
        textbook(QCQI)          & 18          & 18        &    1.02        &    18      & 18          &   0.82      \\
        CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
        % others                  & 7          &  6         &   77.20    &    7        &   6     &  2.13     \\
        % circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
        % research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
        labelled Dirac notation      &   -         &   -          &     -           &      -      &        18       &     22.72    \\
        \hline
    \end{tabular}        
}

@@

\subsubsection{CoqQ}
As to expressibility, D-Hammer encodes 158 examples for the CoqQ part, 4 less than DiracDec. This is because the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$. We found this feature is almost not used and removed the support. For decidability, we improved the rewriting rules about sum for D-Hammer, and proved several more examples failed by DiracDec. 
\Cref{fig: CoqQ plot} shows a direct comparison of their time efficiency.
We can observe that for small examples, D-Hammer is slower because of the marginal overhead. 
When the running time goes up, D-Hammer provides 2 to 40 times of acceleration in average.
The relationship between efficiency improvement and example structure further supports the necessity of our algorithm. 
We find that examples with great efficiency improvement, e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use deeply nested summations, where our algorithm has an effect.

>>>>

\section{Implementation and Case Study}
The improvement and extensions above concludes in our implementation called D-Hammer, a solver written in \CC. It has a parser built by ANTLR4, and scalar reasonings are powered by a Mathematica kernel. The user can use commands to make definitions and assumptions in the maintained context, conduct the normalization and equivalence checking, and obtain the rewriting trace output. 
It can be used from the command line interactively, or can be integrated into other \CC\ projects as a library.
This implementation is tested on the benchmark of the previous work and new examples for labelled Dirac notation, demonstrating significant improvement in expressivity and efficiency. 
The tool and example encodings will be open-source and public available.

@@

\subsubsection{Use Case}
As a tutorial, we encode the motivating~\Cref{ex: motivating}, examine and explain how to check it using D-Hammer. The encoding is shown below.

@@

\begin{center}
    \begin{tabular}{c|c c c|c c c}
        \hline
        \multirow{2}{*}{source} & \multicolumn{3}{c|}{DiracDec} & \multicolumn{3}{c}{D-Hammer} \\
        \cline{2-7}
                                 & expressable & success & time(s)           & expressable & success & time(s)                 \\
        \hline
        textbook(QCQI)          & 18          & 18        &    1.02        &    18      & 18          &   0.82      \\
        CoqQ                    & 162          & 156       &    48.69       &   158     &  158   &     9.74     \\
        % others                  & 7          &  6         &   77.20    &    7        &   6     &  2.13     \\
        % circuits                 & 2          & 2       &    17.67       &   3     &  2   &     1.4     \\
        % research paper                & 4          & 4         &  59.53       &   4    & 4       &  0.73     \\
        % labelled Dirac notation      &   -         &   -          &     -           &      -      &        18       &     22.72    \\
        \hline
    \end{tabular}        
\end{center}

@@

\subsubsection{CoqQ}
As to expressibility, D-Hammer encodes 158 examples for the CoqQ part, 4 less than DiracDec. This is because the language for DiracDec has the support for projectors $\texttt{fst}$ and $\texttt{snd}$ on basis pairs, satisfying $\texttt{fst} (s, t) = s$ and $\texttt{snd} (s, t) = t$. We found this feature is almost not used and removed the support. For decidability, we improved the rewriting rules about sum for D-Hammer, and proved several more examples failed by DiracDec. 
\Cref{fig: CoqQ plot} shows a direct comparison of their time efficiency.
We can observe that for small examples, D-Hammer is slower because of the marginal overhead. 
When the running time goes up, D-Hammer provides 2 to 40 times of acceleration in average.
The relationship between efficiency improvement and example characteristics further supports the necessity of our algorithm. 
We find that examples with great efficiency improvement, e.g. COQQ-129 and COQQ-148 shown in~\Cref{fig: CoqQ plot}, tend to use deeply nested summations, where our algorithm has an effect.
